Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-07-23 exp1/iswc2013.tex
Describe fact designs; normal plots
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@611 43ef1190-90fd-4dd7-a769-d0121f02b80e

In this paper, we employ techniques from the well-established field in
statistics known as Design of Experiments (DoE) to point out a
direction that promises to address these problems. First, we design a
simple 8-parameter experiment, and then we demonstrate how a full
factorial experiment can meaningfully provide a summary of the
statistics.

Then, we demonstrate how a fractional factorial experiment can be used
to maintain a structured approach to the problem while reducing the
number of runs the experiment needs, and we discuss the trade-offs
involved. We then continue to discuss techniques that show our
simplistic experiment to be flawed, yet scientifically better than the
practice of benchmarking. Finally, we outline a road-map to establish
the use of these techniques in SPARQL endpoint evaluations both as a
practical tool and as a scientifically more rigorous methodology.

\subsection{Design of Experiments Concepts}

@@

% TODO: 
% 1) what are effects
%% 2) what are levels
%% 3) what are factors 
%% 4) Can run 2^8 experiments, call full factorial
%% 5) What are normal plots
%% 6) explain interaction effects

@@

The above experimental setup is executed across a design matrix
generated with the DoE.base package, which returns a data frame
consisting of the factors with a column containing the corresponding
curl measurements. To analyze the experiment, we start by fitting a
linear model to the data with all factors and their interactions. It
is shown in \cite{wu2009experiments} that the effect, i.e. the change
from level 1 to 2 is twice the coefficients of the fit. Then, we
generate a normal plot, see Figure~\ref{fig:fullnormal}. In this plot,
if the null hypothesis $H_0$ holds, all estimated effects will fall on
a straight line, as their means would be zero. Any significant
departure from a straight line can be interpreted as a significant
effect (see Section~4.8 of \cite{wu2009experiments} for a detailed
explanation). In our case, it is most important to note that any
effect that is negative means the effect enhances the performance of
the SPARQL endpoint, the runtime decreases. If ``Implement'' and its
interactions are negative conditional on that the factors involved
with its interactions are on their high level, then it supports the
alternative hypothesis, i.e. we have successfully improved the
performance.

>>>>

In this paper, we employ techniques from the well-established field in
statistics known as Design of Experiments (DoE) to point out a
direction that promises to address these problems. First, we design a
simple 8-parameter experiment, and then we detail an experiment that
can meaningfully provide a summary of the statistics.

Then, we demonstrate how the number of runs required in the experiment
can be reduced while maintaining a structured approach to the problem,
and we discuss the trade-offs involved. We then continue to discuss
techniques that show our simplistic experiment to be flawed, yet
scientifically better than the practice of benchmarking. Finally, we
outline a road-map to establish the use of these techniques in SPARQL
endpoint evaluations both as a practical tool and as a scientifically
more rigorous methodology.

\subsection{Key concepts of Design of Experiments}

@@

In an experiment, there are several factors, and one may run the
experiment by measuring the response variable for every combination of
levels. In a two-level experiment, this will result in $2^n$ runs,
where $n$ is the number of factors. This is called a \emph{full
  factorial experiment}. This is done in Section~\ref{sec:full}.

In many fields of science, experimental economy is extremely
important, and so, DoE offers extensive methodology to run a certain
fraction of the runs at the price of explanatory power. Such
experiments are called \emph{fractional factorial experiments} and is
covered in Section~\ref{sec:frac}.

After measurements are done, we seek to estimate \emph{effects}. To
see how the \emph{main effects} can be estimated, consider a factor
$A$. We proceed to find the averages $a_1$ and $a_2$ of all measurements at level~1 and
on level~2 respectively. Then the main effect of $A$ is $a_2 - a_1$.

Similarly, \emph{interaction effects} for 2 factors $AB$ are defined
by comparing averages for equal versus non-equal levels of $A$ and
$B$.

This can be generalized to larger combinations of factors, but in
practice, this analysis is done by fitting a linear model to the data
with all factors and their interactions. It is shown in
\cite{wu2009experiments} that the effect is twice the coefficients of
the fit.

The next step is to understand which factors are important or
significant.  One approach is to plot the sorted effects against the
normal distribution. To understand why, note that if there is nothing
of interest, the measurements have a certain normal distribution
purely due to noise. When plotted against a normal distribution, the
plot would be a straight line. If there are any deviations from the
straight line, it implies that they may be significant. Such plots can
be found in
figures~\ref{fig:fullnormal},~\ref{fig:frac32normal}~and~\ref{fig:frac64normal}. The
range in the $y$-axis is dictated by the number of runs.

@@

The above experimental setup is executed across a design matrix
generated with the DoE.base package, which returns a data frame
consisting of the factors with a column containing the corresponding
curl measurements. Then, we
generate a normal plot, see Figure~\ref{fig:fullnormal}. In this plot,
if the null hypothesis $H_0$ holds, all estimated effects will fall on
a straight line, as their means would be zero. Any significant
departure from a straight line can be interpreted as a significant
effect (see Section~4.8 of \cite{wu2009experiments} for a detailed
explanation). In our case, it is most important to note that any
effect that is negative means the effect enhances the performance of
the SPARQL endpoint, the runtime decreases. If ``Implement'' and its
interactions are negative conditional on that the factors involved
with its interactions are on their high level, then it supports the
alternative hypothesis, i.e. we have successfully improved the
performance.
